---
title: "MATH 216 Homework 1"
author: "Andrea Narciso"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---

```{r, echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr) 
library(lubridate)

# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>% 
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")

# NOTE: THIS WILL PUT THE OUTPUT SHOW UP IN THE RMARKDOWN DOC VS CONSOLE
# Tools > global options > R Markdown < check "show output line for all Rmarkdown docs"
# can also select "show output preview in... <check dropdown>" 
```





## Admistrative:

Please indicate


* Who you collaborated with: None
* Roughly how much time you spent on this HW so far: 8 hours
* The URL of the RPubs published URL [here](http://rpubs.com/anarciso/HW-1).
* What gave you the most trouble: Trying to figure out what combination of commands to use and how to solve an ambiguous problem. Trying to solve the errors produced when knitting the file. 
* Any comments you have:





## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.

Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

# # Calculates the total number of flights per day (denom)
# flights_per_day <- flights %>% 
#   group_by(date) %>% 
#   tally() %>% 
#   rename(num_flights_daily = n)
# # flights_per_day
# 
# # Number of flights per day delayed for over 30 mins
# delay_gt_30 <- flights %>% 
#   filter(arr_delay > 30) %>% 
#   group_by(date) %>%
#   tally() %>% 
#   rename(num_delayed_daily = n)
# # delay_gt_30
# 
# # Calculates proportion of flights delayed >30 mins per day
# prop_delay <- inner_join(flights_per_day, delay_gt_30, by="date") %>% 
#   mutate(prop = round(num_delayed_daily/num_flights_daily, 2))
# # prop_delay

# Calculates proportion of flights delayed >30 mins per day (CONDENSED)
prop_delay <- flights %>% 
  filter(!is.na(dep_delay)) %>% 
  mutate(over30 = ifelse(dep_delay>30, TRUE, FALSE)) %>% 
  group_by(date) %>% 
  summarise(prop=mean(over30))
# NOTE: THIS USES AN IFELSE STATEMENT THAT DRAMATICALLY SHORTENS THE CODE


# Graphs date vs prop
date_v_prop_plot <- ggplot(data=prop_delay, aes(x=date, y=prop)) + 
  geom_point() + 
  geom_smooth(se=FALSE) + 
  labs(x="Date",
     y="Proportion of Flights Delayed > 30 Minutes",
     title="Proportion of Flights Delayed by Date")
date_v_prop_plot

```

The graph above represents a time series of the proportion of flights delayed for more than 30 minutes in the year 2011. It shows that the highest proportion of flights are delayed in the summer, from May to August.



## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?

```{r, echo=FALSE, fig.width=12, fig.height=6}

# Carriers and Plane Models
carrier_model <- flights %>% 
  select(plane, carrier)
# carrier_model

# Year and Plane Models
year_model <- planes %>% 
  select(plane, year)
# year_model

# Carrier and Plane Models by Decade
carrier_decade <- inner_join(carrier_model, year_model, by="plane") %>% 
  filter(!(is.na(year))) %>% 
  mutate(decade_produced = floor(year/10)*10) %>% 
  mutate(carrier = as.factor(carrier), 
         decade_produced = as.factor(decade_produced)) %>% 
  group_by(decade_produced, carrier) %>% 
  tally() %>% 
  rename(num_per_decade = n) 
# carrier_decade

#Plots
carrier_decade_plot <- ggplot(data = carrier_decade, 
                              aes(x = carrier, y = num_per_decade, 
                                  fill = decade_produced)) + 
  geom_bar(stat="identity", position="stack") +
  labs(title = "Number of Planes Produced by Decade for Each Carrier",
       x = "Carrier", 
       y = "Number of Planes Produced")
carrier_decade_plot

# what if try position = "fill"???

```

The above graph presents the number and proportion of planes produced by decade for each carrier. It shows that the major carriers--which I define as the carriers with the largest fleet-- include CO, OO, WN and XE. These major carriers predominantly own planes produced in the 1990s and the 2000s. 

ANALYSIS?????


## Question 3:

* What states did Southwest Airlines' (WN) **flight paths** (y/n) tend to fly to?
* What states did Southwest Airlines' **flights** (number of existing flights) tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

# NOTE: flight paths = highest number of distinct flights to locations w/in a state
#       thus, number of there-and-back,  ==doesn't matter
# NOTE: flights = highest number of total flights to the different lications w/in a state

# Combines the flights and airports data
# adding the information of the airport to each flight

flights_airports <- left_join(flights, airports, by=c("dest"="iata"))
# flights_airports

# Adds the region information to each flight

flights_airports_region <- left_join(flights_airports, states, by="state")
# flights_airports_region

# Narrows down the flights data to only southwest flights,
# Changes the destination to type="factor", and removes irrelevant data

southwest_flights <- flights_airports %>% 
  filter(carrier == "WN", !(is.na(state))) %>% 
  select(date, carrier, flight, dest, airport, state) %>% 
  mutate(dest = as.factor(dest))
# southwest_flights

# 1. FLIGHT PATHS 

# Tallied the number of distinct flights, destinations and states
SW_flight_paths <- southwest_flights %>% 
  group_by(state, dest, flight) %>% 
  tally() %>% 
  rename(num_flights = n)
# SW_flight_paths

# Tallied the number of distinct flights and destinations by state
SW_flight_paths_by_state <- SW_flight_paths %>% 
  group_by(state) %>% 
  tally() %>% 
  arrange(desc(n)) %>% 
  head(n = 5) %>% 
  rename("Number of Flight Paths" = n,
         "State" = state) %>% 
  knitr::kable()
SW_flight_paths_by_state

#NOTE: group_by() THEN tally()


# 2. FLIGHTS

# Counts the number of flights per state
# and sorts the states in increasing order

SW_flights_by_state <- southwest_flights  %>% 
  count(state) %>% 
  rename(num_flights = n) %>% 
  arrange(desc(num_flights)) %>% 
  head(n = 5) %>% 
  rename("Number of Flights" = num_flights,
         "State" = state) %>% 
  knitr::kable()
SW_flights_by_state

# NOTE: sort using arrange(), not sort
# NOTE: use head(n= num_rows) to get TOP 5/10
# NOTE: count() does both group_by() and tally()

```

The first table provides the top 5 states that Southwest Airlines' flight paths tend to fly to. In the original data, one flight number had multiple destinations, which can be attributed to the existence of connecting flights. Consumers are able to take individual flights within a longer, connecting flight; thus, these consumers would likely consider this flight as a "flight path" of the carrier. Therefore, I treated each flight within a connecting flight as a distinct flight path. 

The second table provides the top 5 states that Southwest Airlines' flights tend to fly to. 




## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to, from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}

# Condenses information from flights and airports data tables into 1 table
# containing date, carrier, flight, plane, dest, airport, city, state
# and filters to only contain flights in the month of July

flights_airports_july <- left_join(flights, airports, by=c("dest"="iata")) %>% 
  # mutate(month = ) %>% 
  filter(month(ymd(date)) == 7) %>% 
  select(date, carrier, dest, airport, city, state)
# flights_airports_july

#Adds the region information to the destination location of each flight

flights_region_july <- left_join(flights_airports_july, states, by="state") %>% 
  filter(!(is.na(region)))
# flights_region_july

# Finds the proportion of flights per region
flights_region_july_prop <- flights_region_july %>% 
  group_by(carrier, region) %>% 
  tally() %>% 
  rename(num_by_region = n) %>% 
  group_by(carrier) %>% 
  mutate(total = sum(num_by_region), prop = round(num_by_region/total, 2)) %>% 
  select(region, prop) %>% 
  arrange(carrier, region) %>% 
  rename("Proportion of Flights" = prop, "Carrier" = carrier, "Region" = region) %>%
  knitr::kable()
flights_region_july_prop

# HOW TO USE THIS HERE????????
# prop_delay <- flights %>% 
#   filter(!is.na(dep_delay)) %>% 
#   mutate(over30 = ifelse(dep_delay>30, TRUE, FALSE)) %>% 
#   group_by(date) %>% 
#   summarise(prop=mean(over30))

```
The data table above shows the proportion of flights going to each region from the IAH airport in Houston. 

